MLH1 Results Outline

### 2020-02-26 doc/Results.rmd
cleaning up english

Of the pooled dataset of `r total.biv.all` single bivalents, 96% of bivalents had either one or two crossovers (Figure X). <these results fit with previous observations on the number of crossovers per chromosome (Stapley 2019)>. The proportion of 1CO:2CO distinguishes the high and low rec males (Figure X). Confirming the intuitive interpertation that high recombination strains have more 2CO bivalents at the expense of 1CO bivalents. Across the female observations the propotions of chromsome classes are not notably different. Most of the low rec strains range from `r min(chrm.prop.male.low$n2CO / chrm.prop.male.low$nbiv)` _musculus^KAZ^_ to `r max(chrm.prop.male.low$n2CO / chrm.prop.male.low$nbiv)` ( _molossinus^MOLF^_ ). In the high rec strains the 2CO proportions are `r chrm.prop.male.high1$n2CO[chrm.prop.male.high1$strain == "SKIVE"] / chrm.prop.male.high1$nbiv[chrm.prop.male.high1$strain == "SKIVE"]` ( _musculus^SKIVE^_ ), `r chrm.prop.male.high1$n2CO[chrm.prop.male.high1$strain == "PWD"] / chrm.prop.male.high1$nbiv[chrm.prop.male.high1$strain == "PWD"]` ( _musculus^PWD^_ ), and `r chrm.prop.male.high1$n2CO[chrm.prop.male.high1$strain == "MSM"] / chrm.prop.male.high1$nbiv[chrm.prop.male.high1$strain == "MSM"]` (_molossinus^MSM^_ ).

While the proportions are significantly different, PWD male and MSM male proportions are more similar to each other than strains from the same subspcies ( _musculus^PWD^_ and _musculus^KAZ^_ chi-square test; 1CO p =`r PWD.KAZ.1CO.prop.test$p.value`, 2CO p=`r PWD.KAZ.2CO.prop.test$p.value`) and _molossinus^MSM^_ and _molossinus^MOLF^_ chi-square test; 1CO p = `r MSM.MOLF.1CO.prop.test$p.value`, 2CO p=`r MSM.MOLF.2CO.prop.test$p.value`).

These results confirm the intuitive conclusion that the rapid male specific evolution in gwRR proceeds though the chromosome level/though increasing the number of 2CO bivalents at the expense of 1CO bivalents.

<(Also comparing the low group together ... Six strain have been designated low rec groups, WSB, LEW, G, KAZ, CZECH, and MOLF.
They have significantly different proportions for 1CO and 2CO (chi-square; 1CO p = `r Low_male.1CO.prop.test$p.value` and 2CO p= `r Low_male.2CO.prop.test$p.value`). This significant p value indicates that at least one of the six strains had significant differences).>

@@

-- Confirmation that it's the increase in 2CO bivalents -- that comprises a higher cell wide number of COs (total number)

Let's look at the Rec landscape --

In order to better deconstruct the genome wide recombiation rate, we decided to deconstruct cell wide average by examining features of the recombiantion landscape at the single bivalent level. recombination landscape refers to the i) total length of the SC-AE (potential area for COs), ii) the position of single crossovers along bivalents and iii) the spacing between multiple COs on the same bivalent.

From our total set of cell images `r length(unique(new.Curate.FULL.org %>% filter(SC.pass == "1"))$fileName )` chromosome objects were isolated by the image analysis software. After the human curation step, `r length(unique(new.Curate.FULL.org %>% filter(hand.foci.count != ""))$Obj.ID )` single bivalent measures were left. (make sure these are the hand.foci bivalents).

@@

The ranges of of rates of isolating chromosomes per cell are `r min(algorithm.stat.category$mean_pass.rate)` in _molossinus^MSM^_ male to 
`r max(algorithm.stat.category$mean_pass.rate)` ) _musculus^KAZ^_ female, we assume that the isolation process is not biased. Because there are hundreds of observations per category, we assume that each of the 19 or 20 chromosomes are equally represented in male and female single bivalent data datasets. We assume the bivalent datasets across all categories are equally eqivilent for representing general patterns.

<curation, x from curation, XY excluded, replication to hand measures>

Using this data set, We address two main questions 1) Which traits are sexually dimorphic? and 2) which traits fit distinsutinguish the high and low recombining strains for males?

<These questions use different datasets: Q1 uses a dataset with only sex matched strains  and Q2 uses the data set with all male observations with including strains including those not in the Q1 set.>
<The same basic models from the MLH1 counts per cell will be used. In addition to basic t.tests and logistic regression models for Q2 to distinguish betwee high and low recombining strains.>

# Q1 Sex Differences in SC-AE Lengths

<!-- breif background on longer female SC  (mouse, human) Gruhn-->

Our data set provides an opportunity to test if the canonical pattern of females having higher gwRR and more SC area) have been uncoupled in the instance of rapid male specific evolution in the gwRR of _musculus^PWD^_  and _molossinus^MSM^_  A simple model might predict if SC area is the strongest predictor for gwRR / CO number, the high recombining males would have greater SC area proportional to the difference in CO number per cell.

<Comparing the means of single bivalents is complicated by the fact we can't identify the chromosome identitied the physical size (in terms of SC-AE) is the closet approximation we can make.>

@@

We use a reduced bivalent data set made up of chromsomes (below the 2nd quartile for SC length of single cells) to make single bivalent groups which exclude the XX homologs from the female data. A total of `r sum(Q1.real.short_mouse_table$n_obs)` 'short' bivalents were isolated from `r sum(Q1.real.short_mouse_table$ncells)` oocytes and spermatocytes.

<Even though this smaller data set has decreased power due to redcued sample size, it offers a mors comparable set of SC-AE meterics across the sexes>

@@

<in the above boxplots of mouse means for shortbivalents>

For all but one strain, _musculus^SKIVE^_, the mouse mean for the short bivalents are significantly longer <how much longer, the largest difference across sexes by strain are>  (t.test; p = `r Q1_mus.av_shortSC_t.test.WSB$p.value`, p =`r format(Q1_mus.av_shortSC_t.test.G$p.value, scientific = F)`, and p=`r format(Q1_mus.av_shortSC_t.test.LEW$p.value, scientific = F)`
for _domesticus^WSB^_, _domesticus^G^_ and _domesticus^LEW^_ respectively.
This pattern is also true for the musculus strins (t.test; p= `r format(Q1_mus.av_shortSC_t.test.PWD$p.value, scientific = F)`, `r Q1_mus.av_shortSC_t.test.KAZ$p.value`, and `r  format(Q1_mus.av_shortSC_t.test.SKIVE$p.value, scientific = F)` for _musculus^PWD^_ , _musculus^KAZ^_ , _musculus^SKIVE^_  respectively.

<EXPLAIN WHY USING TOTAL SC AREA>

@@

# Q1. 1CO Position Sex Differences

>>>>

Of the pooled dataset of `r total.biv.all` single bivalents, 96% of bivalents had either one or two crossovers (Figure X) with proportion of 1CO:2CO distinguishes the high and low rec males (Figure X). THis confirms the intuitive interpertation that high recombination strains are enriched for 2CO bivalents at the expense of 1CO bivalents. Across the female observations the proportions of chromosome classes are not notably different.

<Most of the low rec strains range from `r min(chrm.prop.male.low$n2CO / chrm.prop.male.low$nbiv)` _musculus^KAZ^_ to `r max(chrm.prop.male.low$n2CO / chrm.prop.male.low$nbiv)` ( _molossinus^MOLF^_ ).>

In the high rec strains the 2CO proportions are `r chrm.prop.male.high1$n2CO[chrm.prop.male.high1$strain == "SKIVE"] / chrm.prop.male.high1$nbiv[chrm.prop.male.high1$strain == "SKIVE"]` ( _musculus^SKIVE^_ ), `r chrm.prop.male.high1$n2CO[chrm.prop.male.high1$strain == "PWD"] / chrm.prop.male.high1$nbiv[chrm.prop.male.high1$strain == "PWD"]` ( _musculus^PWD^_ ), and `r chrm.prop.male.high1$n2CO[chrm.prop.male.high1$strain == "MSM"] / chrm.prop.male.high1$nbiv[chrm.prop.male.high1$strain == "MSM"]` (_molossinus^MSM^_ ).

While the proportions are significantly different
(p = `r`), PWD male and MSM male proportions are more similar to each other than strains from the same subspcies ( _musculus^PWD^_ and _musculus^KAZ^_ chi-square test; 2CO p=`r PWD.KAZ.2CO.prop.test$p.value`) and _molossinus^MSM^_ and _molossinus^MOLF^_ chi-square test; 2CO p=`r MSM.MOLF.2CO.prop.test$p.value`).

@@

In order to deconstruct the cell wide MLH1 counts
deconstruct we 
look at the recombination landscape at the single bivalent level.

From our total set of cell images `r length(unique(new.Curate.FULL.org %>% filter(SC.pass == "1"))$fileName )` chromosome objects were isolated by the image analysis software. After the human curation step, `r length(unique(new.Curate.FULL.org %>% filter(hand.foci.count != ""))$Obj.ID )` single bivalent measures were left.

We use these observations to look at the recombination landscapes, here refereing to the i) total length of the SC ii) the position of single crossovers along bivalents and iii) the spacing between multiple COs on the same bivalent. <the position is general, not genomic corrdinates>

@@

A potential downside of the image analysis algorithm we used is that not all bivalents per cell are isolated. The ranges of rates of isolating chromosomes per cell are `r min(algorithm.stat.category$mean_pass.rate)` in _molossinus^MSM^_ male to 
`r max(algorithm.stat.category$mean_pass.rate)` ) _musculus^KAZ^_ female, we assume that the isolation process is not biased. However, because there are hundreds of observations per category, we assume the set of single bivalents are equally representative of general patterns.

Using this data set, we address two main questions 1) Which traits are sexually dimorphic? and 2) Which traits fit distinguish the high and low recombining strains for males?

# Q1 Sex Differences in SC-AE Lengths
<!-- breif background on longer female SC  (mouse, human) Gruhn-->

(in many mammal species -- SC lengths are longer in females - yet most all of these observations come from female biased heterochiasmy species) -- Our data set provides an opportunity to test (how SC length and chromatin compaction relates to heterochiasmy)

if the canonical female biased heterochiasmy pattern and long SC lengths are uncoupled in instances / our strain with male-biased heterochiasmy ( _musculus^PWD^_  and _molossinus^MSM^_). A simple model might predict if SC length is the strongest predictor for number of MLH1 foci er cell the high recombining males would have longer SC length proportional to differences in the number of MLH1 foci per cell.

@@

To account of the obvious confounding effect of sex chromosome differences in females and males, we use a reduced bivalent data set made up of chromsomes (below the 2nd quartile for SC length of single cells) to make single bivalent groups which exclude the XX homologs from the female data. A total of `r sum(Q1.real.short_mouse_table$n_obs)` 'short' bivalents were isolated from `r sum(Q1.real.short_mouse_table$ncells)` oocytes and spermatocytes. Even though this smaller data set has decreased power due to redcued sample size, it offers a mors comparable set of SC-AE meterics across the sexes.

@@

For all but one strain, _musculus^SKIVE^_, the mouse mean for the short bivalents are significantly longer <how much longer the largest difference across sexes by strain are>  
**`r` the magnitude differences between the mouse mean short bivalent SC range from X to X**

(t.test; p = `r Q1_mus.av_shortSC_t.test.WSB$p.value`, p =`r format(Q1_mus.av_shortSC_t.test.G$p.value, scientific = F)`, and p=`r format(Q1_mus.av_shortSC_t.test.LEW$p.value, scientific = F)` for _domesticus^WSB^_, _domesticus^G^_ and _domesticus^LEW^_ respectively.
This pattern is also true for the musculus strins (t.test; p= `r format(Q1_mus.av_shortSC_t.test.PWD$p.value, scientific = F)`, `r Q1_mus.av_shortSC_t.test.KAZ$p.value`, and `r  format(Q1_mus.av_shortSC_t.test.SKIVE$p.value, scientific = F)` for _musculus^PWD^_ , _musculus^KAZ^_ , _musculus^SKIVE^_  respectively.

@@

#Q1. 1CO Position Sex Differences
